\section{Fuzzy rho}

\subsection{Manifest of goods}

We need an action on process states, together with a binary operation
taking a pair of names to an element of that action.

\begin{mathpar}
  ( - * - ) : A \times \mathsf{Proc} \rightarrow \mathsf{Proc} \\
  ( - \circ - ) : \mathsf{@Proc} \times \mathsf{@Proc} \rightarrow A
\end{mathpar}

Equipped with these we can define a new $\mathsf{COMM}$ rule

\begin{mathpar}
  %for( y <- x )P | u!( Q ) -> P{ @(( u o x )*Q)/y }
  \inferrule* [lab=COMM] {} {\mathsf{for}( y \leftarrow x )P \;\mathsf{|}\; u!(\arrvec{Q})
    \red P\substn{\quotep{(u \circ x)*Q}}{y}} \\
\end{mathpar}

The intuition is that communication isn't based on exact
match. Instead, the strength of the match, as measured by $u \circ x$,
determines how much of the data is transmitted, which is calculated by
$(u \circ x)*Q$.

To tighten these intuitions we ask for a ``measure'' on
$\mathsf{Proc}$. We use an order-theoretic gadget, namely a quantale,
to compare process states. That is, we ask for $\mu : \mathsf{Proc} ->
\mathsf{K}$ where $\mathsf{K}$ is a quantale. Finally, we require the
action to carry the order information. That is $A = \mathsf{K}.$ Then
we need that

\begin{mathpar}
  a1 \leq a2 \Rightarrow \mu( a1*P ) \leq \mu( a2*P )
\end{mathpar}

Under these conditions we can calculate

\begin{mathpar}
  %for( y <- u1 )P | x1!( Q ) -> P{ @(( u1 o x1 )*Q)/y }
  \inferrule* {} {\mathsf{for}( y \leftarrow x )P \;\mathsf{|}\; u_{1}!(\arrvec{Q})
    \red P\substn{\quotep{(u_{1} \circ x)*Q}}{y}} \\
  %for( y <- u2 )P | x2!( Q ) -> P{ @(( u2 o x2 )*Q)/y }
  \inferrule* {} {\mathsf{for}( y \leftarrow x )P \;\mathsf{|}\; u_{2}!(\arrvec{Q})
    \red P\substn{\quotep{(u_{2} \circ x)*Q}}{y}} \\
\end{mathpar}

Which ensures that if the match is worse, less data is transmitted.

\subsection{A model}

First, let's adopt a notation that when $x = \mathsf{@}P$, then $\mathsf{\#}x = P$. Now, we set $A = \mathsf{@Proc}$. This choice affords the following definitions.

\begin{mathpar}
  x_{1} \circ x_{2} = @( \mathsf{\#}x_{1} | \mathsf{\#}x_{2} ) \\
  u * P = P\{ (u \circ x)/x | x \in \mathsf{FN}( P ) \} \\
  \mu( P ) = \mathsf{@}P
\end{mathpar}

which the reader can check satisfies the conditions.
