\section{Introduction}\label{sec:introduction} % (fold)

One of the notable things about the Einstein field equations
\cite{Misner1973} is their mutual recursion. The stress energy tensor
describing the distribution of matter and energy informs the metric
tensor, which in turn informs the stress energy tensor. Any red
blooded computer scientist sees this and immediately wants to express
it as a couple of mutually recursive functions and/or data structures.

Meanwhile, in computer science a revolution in the notion of location
has happened and all but gone unnoticed. On the one hand, Huet's
zipper \cite{DBLP:journals/jfp/Huet97} generalizes the intuitions of
the Dedekind cut for all differentiable data structures. Specifically,
for a data structure described by a differentiable functor, $T$, an
element of the data type $\partial T \times T$ describes a location
in an element of $T$ \cite{DBLP:conf/popl/McBride08}.

On the other, Milner's $\pi$-calculus transforms the
$\lambda$-calculus by reifying the site of interaction as a channel,
and introducing parallel composition \cite{milner91polyadicpi}. This
change simultaneously gives computations autonomy, allowing them to
roam far outside the bounds of mere textual juxtaposition, and yet a
means of finding each other to effect data exchange and computational
evolution. Meredith's rho-calculus takes Milner's insight a step
further \cite{DBLP:journals/entcs/MeredithR05}, making the reified
site of interaction into a fully first class entity with a complexity
of structure equivalent to computations.

A question worth exploring is whether these two notions of location
can be combined, yielding a site of interaction corresponding to a
location in a computation. The thought is that such a reconciliation
of the two ideas would allow for a computational calculus enjoying
something like the intuitions underlying the mutually recursive
dynamics in Einstein's field equations. Specifically, locations say
where computation is distributed, computation changes where things are
located.

In this paper we describe a calculus that does effect a reconciliation
of the two notions of location. It provides a sandbox for exploring a
wide range of discrete space-like phenomena in a fairly traditional,
compact and tractable calculus. In particular, it enjoys a notion of
bisimulation, allowing one to reason equationally about computational
behavior, and by extension about spatial phenomena.

\subsection{Summary of contributions and outline of paper}

\subsubsection{Summary of contributions}
We present a calculus, dubbed the space calculus (from outer space!)
whose primitives are channels built from locations in computations
reified as data structures. This gives us the ability to model a
variety of space-like phenomena. One important feature of the calculus
is that space \emph{emerges} from computation. This makes it very
distinct from work like Cardelli and Gardner's processes in space
\cite{DBLP:conf/cie/CardelliG10} where they embed more standard
geometric primitives in a process calculus setting.

\subsubsection{Outline of paper}
We provide the source of inspirations for the calculus and in so doing
provide a more colorful review of related work. Then we present the
calculus and a few examples of calculating with it. We present our
main theorem demonstrating a fully abstract encoding of the ambient
calculus into the space calculus, providing what we believe to be the
first fully abstract encoding of an ambient style calculus into a
substitution-based calculus. Finally, we conclude with some directions
for future work.

% section introduction (end)
